Trade ornaments as indicators of social changes resulting from indirect effects of colonialism in northeastern Taiwan

### 2020-09-25 analysis/paper/paper.Rmd
edit based on proof

---
title: "Ornaments as indicators of social changes resulting from indirect effects of colonialism in northeastern Taiwan"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      #reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/antiquity.csl" # Insert path for the bib-style
abstract: |
  Long-lasting indirect impacts on Indigenous peoples in the periphery of colonial control are poorly understood, especially in East Asia. Trade ornaments from Kiwulan (1350–1850 CE) in northeastern Taiwan show indirect impacts of European colonial activities on local societies. The diversity of ornaments was greater during the period of European presence compared to previous periods, and their spatial distribution was more clustered in the residential area. The change reflects an increased use of ornaments and an uneven distribution of ornaments in a pericolonial context. This may hint at increased social inequality resulting from European colonial influence. Trade ornaments give insights into the increased social inequality stimulated by a European colonial presence, and show the agency of Indigenous people to incorporate ornaments into their social system.
keywords: |
  Historical archaeology; European colonial period; Trade ornaments; Exchange networks; spatial pattern; East Asia; Taiwan
highlights: |
  1. Evidence of the quantity, variety, and spatial distribution of trade ornaments in both local cultural and colonial context is provided.
  2. Results of hypothesis testing point to greater amount and diversity of ornament types with non-random, and most highly concentrated spatial clusters during the European presence.
  3. Trade ornament as a prestige good may result in an increase in social inequality due to competition of wealth accumulation between individuals.
  4. Trade ornament can be a good proxy to detect indirect colonial influence on local Indigenous populations and inform on social changes relating to trade with colonial groups.
---

@@

One of the most commonly traded types of object in this region were ornaments such as glass and stone beads [@Chen2007; @Li2014; @NMTH2005]. Personal adornments in the archaeological record are useful as signal of an individual's status [@Joyce2005; @Scaramelli2005]. The consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000]. In this paper, we explore archaeological ornaments from Kiwulan spanning the pre-European contact period, period after the Spanish and Dutch presence, and the following period after the Chinese presence. We address the question of whether indirect colonial influences on the Indigenous populations can be detected through the ornament assemblages.

# Ornaments in complex exchange network during the late Iron Age and early historical period

The island of Taiwan lies at the junction of mainland China, Southeast Asia, and Northeast Asia in the Pacific Ocean. The prehistory of Taiwan island could be roughly divided into three major periods, Palaeolithic (c. 27,000 BP- 5000BP), Neolithic (c. 6500- 2000BP), and Iron age (c. 2000- 400BP) with slight regional differences in onset of each period and variations in style of artifacts and assemblages [@Chen2017; @Liu2011]. It is generally accepted that Taiwan entered the historical period in the early 17th century due to the colonial activities of the Spanish and the Dutch who played an important role in keeping written records about Taiwan. The European colonial presence in Taiwan ended in 1662 when the Dutch were defeated by the kingdom of Tungning, founded by Koxinga from China. Later in 1683, Taiwan was incorporated into the Qing dynasty in China and a large wave of Han Chinese migrated to Taiwan during the late 18th century. Because of natural safe harbors, northeastern Taiwan was involved in a regional trade network through cross-culture interactions with Chinese merchants since the 14th century, and later the global trade network with the Europeans in the 17th century brought more trade goods circulated in Southeast Asia into Taiwan [@Chen2005; @Liu2017]. Although located on the periphery of regional trade centers, Yilan was connected to trade networks via visits of other Indigenous groups, Chinese merchants, and Europeans, via sea.

The European presence in northern Taiwan started with the Spanish who founded Fort San Salvador at Heping Dao, Keelung in 1626, and Fort San Domingo in 1629 at Tamsui (Figure \@ref(fig:KWL-map)). They sent missionaries to local Indigenous settlements in this region [@Blusse2000, pp. 343] and kept records about their observations of Indigenous communities. A Dominican priest in 1632 reported that the Taparri, an Indigenous tribe from northern Taiwan, exchanged carnelian beads with other Indigenous groups. This form of exchange was widespread and even the Spanish soldiers used carnelian beads as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is further indicated by their role in bride price payments, and compensation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the women shamans in the tribe would use carnelian beads as magical items in ritual healing practices [@Borao2009, pp. 122-151]. Records of an Indigenous funeral describe the use of carnelian beads in ritual contexts, with more carnelian beads, pottery, and cloth placed into the graves of more influential people to indicate their family's higher status [@LiandWu2006, pp. 153]. While a full critical analysis of these historical accounts remains to be produced, we take them to minimally indicate that carnelian beads were already treated as prestige goods in Yilan before the arrival of Europeans. In 1642, the Dutch Vereenigde Oostindische Compagnie (VOC) defeated the Spanish and took over their forts in northern Taiwan. They introduced a feudal system in an attempt to control the Indigenous communities by asking Indigenous leaders to attend an annual ceremony for demonstrating their loyalty and paying tribute [@Andrade2007, ch. 9; @Kang2016, ch. 4]. The Dutch provided beads and other goods based on negotiations with Indigenous communities to secure alliances in the annual ceremony or during their travels [@Kang2016, ch. 6]. We might predict that the activities of the Dutch feudal system to build and maintain alliances resulted in an increase in the amount and diversity of ornaments in northeastern Indigenous communities during this period.

Chinese historical records from 1829, 1837, and 1852 during the Qing dynasty (1616-1911) contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, Indigenous people in Yilan wore ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, fish-shaped necklaces made of metal threads had high value due to their delicacy and the exotic materials invested in production. These were usually possessed by wealthy people. Other people wore carnelian beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonization, an academic field survey for plains Indigenous groups reported that fish-shaped metal necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232]. Although these historical records are fragmentary and may contain some biases [@Galloway2006] that have not yet been studied in detail, we find consistency among multiple sources in their descriptions of how ornaments represent high status or specialized social roles in Indigenous communities in Yilan. Compared to the European period, there are fewer documentary mentions of beads in the Chinese period and the descriptions are limited to clothing, but these generally confirm the role of beads as status markers.

Ornaments found in northeastern Taiwan in the early historical period, including glass beads, stone beads, and metal ornaments, are considered to have been imported from other regions. This is because of a lack of archaeological evidence of beadmaking waste, metalworking, or accessible local raw materials. The chemical composition of glass beads from this region shows a high content of lead and, together with the winding/folding technique, these details suggest a Chinese beadmaking tradition [@Cheng2008; @Gan2006; @Wang2018]. Although there is a wide variety of metal ornaments such as bells, bracelets, rings, and pendants, the common components of metal ornaments are brass and copper, with a small number made from lead and tin that indicates multiple origins that include Southeast Asia [@Chen2011]. There is no direct evidence showing European delivery of beads, however, a large amount of the glass beads containing gold foil (hereafter, gold-foil beads) at Kiwulan might have been introduced by the Spanish through economic activities because similar beads were found at Luzon, northern Philippines, as part of the trading route of the Spanish between 16-19th century [@Wang2007]. Both archaeological evidence and historical records indicate northeastern Taiwan was involved in regional networks with East Asia in the late Iron age. These included Chinese merchants trading metal items, clothes, and beads with local Indigenous people in Taiwan in exchange for local resources. The foreign-made large dark brown glazed stoneware jars frequently found in European shipwrecks were also commonly found from many sites in Taiwan, suggesting direct or indirect interactions. Despite the Chinese origin of some ornaments at Kiwulan, there is compelling evidence that a large amount of ornaments found at 17th century sites resulted from European colonial and economic activities in the region.

```{r, KWL-map, fig.cap = "Map showing the location of Kiwulan, and other places in northern Taiwan named in the text. Map data from naturalearthdata.com", fig.width=image_width_2_col_inch}
library(here)
knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.jpg"))
```

@@

Archaeological ornaments from Kiwulan (Figure \@ref(fig:KWL-map)) come from a rescue archaeology project that was conducted between 2001-2004 in advance of a water diversion project and road bridge construction. The excavations used 2 mm and 1.5 mm mesh screens and covered eight open area sections in total of 262 squares (4 m by 4 m) reaching 3,814 m^2^ [@Chen2007]. The nearly 2 m thick archaeological deposits reveal a large amount of artifacts, burials, middens, post-holes, wooden pillars, and stone structures, all of which indicates it was a long-term settlement. Artifact locations were recorded to the 2 x 2 m sub-square they were recovered in; they lack individual point provenance. Based on the continuity of deposition and the frequency of artifacts, the center of the site is the open area consisting of the A and D sections, which is also the study area where our samples come from (Figure \@ref(fig:KWL-sam-area)). In the AD area, post-holes were found aligned in a north-south direction in intervals with construction marks, which have been interpreted as the remains of stilt house structures. At the north margin of the dwelling place were burials that are mostly oriented in an east-west direction.

@@

```{r, KWL-sam-area, fig.cap = "Map showing the largest section of excavation areas at Kiwulan, and the distribution of forty squares sampled in this paper presented in red with square ID numbers. Small dots represent the locations of post-holes. Each square is 4 x 4 m", fig.width=image_width_2_col_inch}
# add to the map and scale and north arrow. Check it out after knitting 
library(sf)
library(cartography)
#devtools::install_github('oswaldosantos/ggsn')
library(ggsn)
# read in the spatial data
burial <- invisible(st_read(here("analysis", "data", "raw_data", "AD_burial.shp"), quiet = TRUE))
post <- invisible(st_read(here("analysis", "data", "raw_data", "AD_postholes.shp"), quiet = TRUE))
AD <- invisible(st_read(here("analysis", "data", "raw_data", "AD_zone.shp"), quiet = TRUE))
AD_data <- invisible(st_read(here("analysis", "data", "raw_data", "AD_withdata.shp"), quiet = TRUE))
location <- invisible(st_read(here("analysis", "data", "raw_data", "location.shp"), quiet = TRUE))
units_40 <- invisible(read.csv(here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv")))

@@

Studying change over time at a site like Kiwulan has some distinct challenges. First, the archaeological deposits are relatively thin. This makes it difficult for excavators to clearly identify distinct chronological phase boundaries. In thin deposits, site formation processes can quickly blur the boundaries or completely mix deposits from different time periods. Second, the time periods we aim to compare are relatively short, just a few human generations. This limits the usefulness of common chronometric methods such as radiocarbon dating, where the age errors are relatively large compared to the periods we are studying. Most of the previously obtained radiocarbon ages from Kiwulan do not span the time period we are investigating here (Figure \@ref(fig:C14-data) and Table \@ref(tab:C14-table-data)). Third, pericolonial contexts such as Kiwulan can involve non-Western notions and experiences of time. For example, a Western concept of time seeks clear temporal divides, and may create material signature of this, but Indigenous experiences of time may prioritize continuity and continuums of gradual change, and their material culture traces may reflect that intentional experience of time [@Trabert2018; @Scheiber2012]. Despite the relatively short period of the European colonization in Taiwan (38-years-long Dutch occupation and 16-years-long Spanish occupation), the indirect impacts of the Europeans through trade networks could go beyond these temporal boundaries and last longer. This is because the use of foreign goods in many Indigenous societies often goes through processes of negotiations, resistance, and transformation. A static time frame may not capture the processes well [@Scheiber2012]. Fourth, the circulation of foreign goods that we use as chronological indicators might start earlier than the historically documented European presence in Taiwan. Similarly, the use of the foreign goods in Indigenous societies could continue after the end of the European colonization [@Mitchell2010]. This limits the usefulness of specific artifact types as chronological markers, and implies they have an error range similar to radiocarbon ages. Unfortunately, we do not have a way to accurately estimate these errors at Kiwulan. Ideally, we would divide the deposits at Kiwulan into many time slices to investigate the possibility of a gradual versus sudden change, and assess errors in the accuracy of artefacts as temporal markers. However, this would result in many analytical units with no finds, so we have grouped excavation units together to create minimum sample sizes suitable for addressing our research questions. The details about the distribution of the temporal indicators, accompanied with stratigraphy data, radiocarbon dates, and archaeological contexts for each sampled excavation unit are provided in Supplementary Online Materials.

A first step to outlining chronology of Kiwulan is indicated by the stratigraphy. There is an upper component (1350-1850 AD, 600-100 BP) and a lower component (650-1150 AD, 1200-800 BP) based on 32 radiocarbon ages, previously published by @Chen2007, and shown here in Figure \@ref(fig:C14-data) and Table \@ref(tab:C14-table-data). There is a sterile deposit between these two components spanning c. 150 years that may be related to dry weather, according to pollen analysis, leading to site abandonment [@Chen2007; @Lin2015]. These component divisions present the differences in color and texture of the deposit, and the content of artifacts such as pottery types. Whether these two components suggest a countinuity of culture from the same people is still under debate [@Chiu2004; @Chen2007]. We focus on the upper component because this component spans the periods when local residents interacted with the Europeans and the Chinese. In our study area, all excavation squares demonstrate signs of continuous human occupation throughout the upper component. Previous work divided the upper component into six analytical units, spanning from the 14th century to the 19th century, according to the types of chronologically diagnostic ceramics, excavation depth, consistency of contexts, and radiocarbon dates [@Hsieh2009; @Wang2011]. To achieve sample sizes suitable for exploring changes in ornament assemblages over time associated with foreign impacts, we assigned the original excavation levels into three phases: before the European contact, after the European presence, and the presence of large Chinese immigrants.

@@

We made refinements of the original chronology to assign contexts into the pre-European, European, and the Chinese periods with six steps of assigning and evaluations using copies of original excavation records and fieldwork notes. First, for each unit, we determined the artificial layer (mostly 10 cm) with indirect colonial impacts in terms of the higher frequency of time indicators representing the European and the Chinese period. Second, we identified the archaeological context (normally 2-3 layers) where the layer belongs to based on the description of soil color, soil texture, and density of potsherds and charcoal. We assumed that a larger population would be reflected by denser distributions of potsherds and charcoal. Third, we checked if there are radiocarbon ages associated with the context as a cross-validation. Since some radiocarbon ages represent long time ranges, we did not consider the ages could determine a phase by themselves, but only as a cross-validation. Fourth, we explored any post-depositional issues that might affect the deposition of time indicators to ensure an appropriate assignment based on the fieldwork notes and excavation report. Fifth, we cross evaluated all adjacent squares to confirm a consistent and continuous context between them. Sixth, we compared our chronology with six analytic units classified by previous studies to make a final decision [@Hsieh2009; @Wang2011]. More detailed data information is provided in Supplementary Online Materials. After assigning the contexts into three analytical units, pre-European contact, the European presence, and the Chinese presence, we assigned ornaments to one of these units according to their find context. We did not use the ornaments themselves as temporal markers because many of the ornaments have been found throughout long time periods in Taiwan. Through those assigning steps with these validation rules, we believe our chronology is reasonable and suitable for discussing the indirect colonial impacts led by trade networks.

Archaeological indicators of indirect impacts of the Europeans at Kiwulan are the higher frequency of light grey glazed jars, known as "An-ping" jars in China and Taiwan, and large dark brown glazed stoneware jars that were introduced to Taiwan during the early 17th century. Large dark brown glazed stoneware jars may have been made in Southeast Asia, but are frequently found in European shipwrecks from this period as vessels for transporting water, wine or other liquids on long voyages. The earliest evidence of light gray glazed jars in this region has been found among the cargo of the Spanish shipwreck _San Diego_, which sunk in 1600 AD [@Dizon2016; @Hsieh1995]. Southeast China is assumed to be the origin of the light gray glazed jars, however these are commonly found at sites in Taiwan that were associated with European activities, such as the Zeelandia fort site in Tainan [@Wang2007]. The jar shapes found at Kiwulan are typical of those found elsewhere in VOC sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. We cannot be sure of the exact process that brought them to Kiwulan: they might have been directly imported by Europeans, by Chinese merchants, or by Indigenous groups via regional networks in north Taiwan. In any case, the high volume of ceramics transported by Europeans, and their high mobility in the shipping trade played an important role in introducing foreign jars to Taiwan.

Those jars were widely distributed across the site and can serve as indicators, together with the radiocarbon dates, to identify the excavation units associated with the pre-European period and the start of European influence at Kiwulan. In addition to stoneware jars as indicators of European presence, around 300 pieces of locally made clay pipes and a few imported pipes were found at Kiwulan. Smoking is likely to have been introduced by Europeans. This custom was widely adopted in many European countries in the 16th century and spread to other regions through interactions [@Uccar2019]. We found that the presence of pipe bowls in the archaeological record here is consistent with distributions of glazed jar fragments, which are far more numerous and widespread across the site (n = 1685). It should be noted that these temporal indicators might have been introduced before direct European contact by Chinese traders, and this could result in some uncertainty in identifying the start of European phase. However, the archaeological evidence shows that the layers with abundant trade ceramics match the 17th century according to the excavation report and previous studies [@Hsieh2009; @Wang2011]. Thus, we focus on identifying the contexts with high frequencies of those ceramics as indicators the early 17th century.

>>>>

---
title: "Trade ornaments as indicators of social changes resulting from indirect effects of colonialism in northeastern Taiwan"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      #reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/antiquity.csl" # Insert path for the bib-style
abstract: |
  Long-lasting indirect impacts on Indigenous peoples in the periphery of colonial control are poorly understood, especially in East Asia. Trade ornaments from Kiwulan (1350–1850 CE) in northeastern Taiwan show indirect impacts of European colonial activities on local societies. The diversity of ornaments was greater during the period of European presence compared to previous periods, and their spatial distribution was more clustered in the residential area. The change reflects an increased use of ornaments and an uneven distribution of ornaments in a pericolonial context. This may hint at increased social inequality resulting from European colonial influence. Trade ornaments give insights into the increased social inequality stimulated by a European colonial presence, and show the agency of Indigenous people to incorporate ornaments into their social system.
keywords: |
  Historical archaeology; European colonial period; Trade ornaments; Exchange networks; spatial pattern; East Asia; Taiwan
highlights: |
  1. Evidence of the quantity, variety, and spatial distribution of trade ornaments in both local cultural and colonial context is provided.
  2. Results of hypothesis testing point to greater amount and diversity of ornament types with non-random, and most highly concentrated spatial clusters during the European presence.
  3. Trade ornament as a prestige good may result in an increase in social inequality due to competition of wealth accumulation between individuals.
  4. Trade ornament can be a good proxy to detect indirect colonial influence on local Indigenous populations and inform on social changes relating to trade with colonial groups.
---

@@

One of the most commonly traded types of object in this region were ornaments such as glass and stone beads [@Chen2007; @Li2014; @NMTH2005]. Personal adornments in the archaeological record are useful as signal of an individual's status [@Joyce2005; @Scaramelli2005]. The consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000]. In this paper, we explore archaeological ornaments from Kiwulan spanning the pre-European contact period, period after the Spanish and Dutch presence, and the following period after the Chinese presence. We address the question of whether indirect colonial influences on the local Indigenous populations can be detected through the ornament assemblages.

# Ornaments in exchange networks during the late Iron Age and early historical period

The island of Taiwan lies at the junction of mainland China, Southeast Asia, and Northeast Asia in the Pacific Ocean. The prehistory of Taiwan can be roughly divided into three major periods, Palaeolithic (c. 27,000- 5000 BP), Neolithic (c. 6500- 2000 BP), and Iron age (c. 2000- 400 BP) with slight regional differences in the onset of each period and variations in style of artifacts and assemblages [@Chen2017; @Liu2011). It is generally accepted that Taiwan entered the historical period in the early 17th century due to the colonial activities of the Spanish and the Dutch who played an important role in keeping written records about Taiwan. The European colonial presence in Taiwan ended in 1662 when the Dutch were defeated by the kingdom of Tungning, founded by Koxinga from China. Later in 1683, Taiwan was incorporated into the Qing dynasty in China and a large wave of Han Chinese migrated to Taiwan during the late 18th century. Because of natural safe harbors, northeastern Taiwan was involved in a regional trade network through cross-culture interactions with Chinese merchants since the 14th century, and later global trade networks with Europeans in the 17th century brought more trade goods circulating in Southeast Asia into Taiwan [@Chen2005; @Liu2017]. Although located on the periphery of regional trade centers, Yilan was connected by sea to trade networks via visits of other Indigenous groups, Chinese merchants, and Europeans.

The European presence in northern Taiwan started with the Spanish who founded Fort San Salvador in 1626 at Heping Dao, Keelung [@Berrocal2020], and Fort San Domingo in 1629 at Tamsui (Figure \@ref(fig:KWL-map)). They sent missionaries to local Indigenous settlements in this region [@Blusse2000, pp. 343] and kept records about their observations of Indigenous communities. A Dominican priest in 1632 reported that the Taparri, an Indigenous tribe from northern Taiwan, exchanged carnelian beads with other Indigenous groups. This form of exchange was widespread and even the Spanish soldiers used carnelian beads as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is further indicated by their role in bride price payments, and compensation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the women shamans in the tribe would use carnelian beads as magical items in ritual healing practices [@Borao2009, pp. 122-151]. Records of an Indigenous funeral describe the use of carnelian beads in ritual contexts, with more carnelian beads, pottery, and cloth placed into the graves of more influential people to indicate their family's higher status [@LiandWu2006, pp. 153]. While a full critical analysis of these historical accounts remains to be produced, we take them to minimally indicate that carnelian beads were already treated as prestige goods in Yilan before the arrival of Europeans. In 1642, the Dutch Vereenigde Oostindische Compagnie (VOC) defeated the Spanish and took over their forts in northern Taiwan. They introduced a feudal system in an attempt to control the Indigenous communities by asking Indigenous leaders to attend an annual ceremony for demonstrating their loyalty and paying tribute [@Andrade2007, ch. 9; @Kang2016, ch. 4]. The Dutch provided beads and other goods based on negotiations with Indigenous communities to secure alliances in the annual ceremony or during their travels [@Kang2016, ch. 6]. We might predict that the activities of the Dutch feudal system to build and maintain alliances resulted in an increase in the amount and diversity of ornaments in northeastern Indigenous communities during this period.

Chinese historical records from 1829, 1837, and 1852 during the Qing dynasty (1616-1911) contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, Indigenous people in Yilan wore ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, fish-shaped necklaces made of metal threads had high value due to their delicacy and the exotic materials used in production. These were usually possessed by wealthy people. Other people wore carnelian beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of the Japanese colonization of Taiwan, an academic field survey for Indigenous groups reported that fish-shaped metal necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232]. Although these historical records are fragmentary and may contain some biases [@Galloway2006] that have not yet been studied in detail, we find consistency among multiple sources in their descriptions of how ornaments represent high status or specialized social roles in Indigenous communities in Yilan. Compared to the European period, there are fewer documentary mentions of beads in the Chinese period and the descriptions are limited to clothing, but these generally confirm the role of beads as status markers.

Ornaments found in northeastern Taiwan in the early historical period, including glass beads, stone beads, and metal ornaments, are considered to have been imported from other regions. This is because of a lack of archaeological evidence of beadmaking waste, metalworking, or accessible local raw materials. The chemical composition of glass beads from this region shows a high content of lead and, together with the winding/folding technique, these details suggest a Chinese beadmaking tradition [@Cheng2008; @Gan2006; @Wang2018]. Although there is a wide variety of metal ornaments such as bells, bracelets, rings, and pendants, the common components of metal ornaments are brass and copper, with a small number made from lead and tin that indicates multiple origins that include Southeast Asia [@Chen2011]. There is no direct evidence showing European delivery of beads, however, a large amount of the glass beads containing gold foil (hereafter, gold-foil beads) at Kiwulan might have been introduced by the Spanish through economic activities because similar beads were found at Luzon, northern Philippines, as part of the trading route of the Spanish between 16-19th century [@Wang2007]. Both archaeological evidence and historical records indicate northeastern Taiwan was involved in regional networks with East and Southeast Asia in the late Iron age. These included Chinese merchants trading metal items, clothes, and beads with local Indigenous people in Taiwan in exchange for local resources. The foreign-made large dark brown glazed stoneware jars frequently found in European shipwrecks were also commonly found from many sites in Taiwan, suggesting direct or indirect interactions. Despite the Chinese origin of some ornaments at Kiwulan, there is compelling evidence that a large amount of ornaments found at 17th century sites resulted from European colonial and economic activities in the region.

```{r, KWL-map, fig.cap = "Map (from Wang and Marwick, 2020) showing the location of Kiwulan, and other places in northern Taiwan named in the text. Map data from naturalearthdata.com", fig.width=image_width_2_col_inch}
library(here)
knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.jpg"))
```

@@

Archaeological ornaments from Kiwulan (Figure \@ref(fig:KWL-map)) come from a rescue archaeology project that was conducted between 2001 and 2004 in advance of a water diversion project and road bridge construction. The excavations used 2 mm and 1.5 mm mesh screens and covered eight open area sections in total of 262 squares (4 m by 4 m) reaching 3,814 m^2^ [@Chen2007]. The nearly 2 m thick archaeological deposits reveal a large amount of artifacts, burials, middens, post-holes, wooden pillars, and stone structures, all of which indicate it was a long-term settlement. Artefact locations were mostly recorded to the 2 x 2 m sub-square they were recovered in. Except for a handful of finds, artefacts lack individual point provenance information. Based on the continuity of deposition and the frequency of artifacts, the center of the site is the open area consisting of the A and D sections, which is also the study area where our samples come from (Figure \@ref(fig:KWL-sam-area)). In the AD area, post-holes were found aligned in a north-south direction in intervals with construction marks, which have been interpreted as the remains of stilt house structures. At the north margin of the dwelling place were burials that are mostly oriented in an east-west direction.

@@

```{r, KWL-sam-area, fig.cap = "Map (from Wang and Marwick, 2020) showing the largest section of excavation areas at Kiwulan, and the distribution of forty squares sampled in this paper presented in red with square ID numbers. Small dots represent the locations of post-holes. Each square is 4 x 4 m", fig.width=image_width_2_col_inch}
# add to the map and scale and north arrow. Check it out after knitting 
library(sf)
library(cartography)
#devtools::install_github('oswaldosantos/ggsn')
library(ggsn)
# read in the spatial data
burial <- invisible(st_read(here("analysis", "data", "raw_data", "AD_burial.shp"), quiet = TRUE))
post <- invisible(st_read(here("analysis", "data", "raw_data", "AD_postholes.shp"), quiet = TRUE))
AD <- invisible(st_read(here("analysis", "data", "raw_data", "AD_zone.shp"), quiet = TRUE))
AD_data <- invisible(st_read(here("analysis", "data", "raw_data", "AD_withdata.shp"), quiet = TRUE))
location <- invisible(st_read(here("analysis", "data", "raw_data", "location.shp"), quiet = TRUE))
units_40 <- invisible(read.csv(here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv")))

@@

Studying change over time at a site like Kiwulan has some distinct challenges. First, the archaeological deposits are relatively thin. This makes it difficult for excavators to clearly identify distinct chronological phase boundaries. In thin deposits, site formation processes can quickly blur the boundaries or completely mix deposits from different time periods. Second, the time periods we aim to compare are relatively short, just a few human generations. This limits the usefulness of common chronometric methods such as radiocarbon dating, where the age errors are relatively large compared to the periods we are studying. Most of the previously obtained radiocarbon ages from Kiwulan do not span the time period we are investigating here (Figure \@ref(fig:C14-data) and Table \@ref(tab:C14-table-data)). Third, pericolonial contexts such as Kiwulan can involve non-Western notions and experiences of time. For example, a Western concept of time seeks clear temporal divides, and may create material signature of this, but Indigenous experiences of time may prioritize continuity and continuums of gradual change, and their material culture traces may reflect that intentional experience of time [@Trabert2018; @Scheiber2012]. Despite the relatively short period of the European colonization in Taiwan (38-years-long Dutch occupation and 16-years-long Spanish occupation), the indirect impacts of the Europeans through trade networks could go beyond these temporal boundaries and last longer. This is because the use of foreign goods in many Indigenous societies often goes through processes of negotiations, resistance, and transformation. A static time frame may not capture the processes well [@Scheiber2012]. Fourth, the circulation of foreign goods that we use as chronological indicators might start earlier than the historically documented European presence in Taiwan. Similarly, the use of the foreign goods in Indigenous societies could continue after the end of the European colonization [@Mitchell2010]. This limits the usefulness of specific artefact types as chronological markers, and implies they have an error range similar to radiocarbon ages. Unfortunately, we do not have a way to accurately estimate these errors at Kiwulan. Ideally, we would divide the deposits at Kiwulan into many time slices to investigate the possibility of a gradual versus sudden change, and assess errors in the accuracy of artefacts as temporal markers. However, this would result in many analytical units with no finds, so we have grouped excavation units together to create minimum sample sizes suitable for addressing our research questions. The details about the distribution of the temporal indicators, accompanied with stratigraphy data, radiocarbon dates, and archaeological contexts for each sampled excavation unit are provided in Supplementary Online Materials.

A first step to outlining chronology of Kiwulan is indicated by the stratigraphy. There is an upper component (1350-1850 CE, 600-100 BP) and a lower component (650-1150 CE, 1200-800 BP) based on 32 radiocarbon ages, previously published by @Chen2007, and shown here in Figure \@ref(fig:C14-data) and Table \@ref(tab:C14-table-data). There is a sterile deposit between these two components spanning c. 150 years that may be related to dry weather, according to pollen analysis, leading to site abandonment [@Chen2007; @Lin2015]. These component divisions present the differences in the colours and textures of the deposit, and the content of artifacts such as pottery types. Whether these two stratigraphic components suggest a countinuity of culture from the same people is still under debate [@Chiu2004; @Chen2007]. We focus on the upper component because this component spans the periods when local residents interacted with the Europeans and the Chinese. In our study area, all excavation squares demonstrate signs of continuous human occupation throughout the upper component. Previous work divided the upper component into six analytical units, spanning from the 14th century to the 19th century, according to the types of chronologically diagnostic ceramics, excavation depth, consistency of contexts, and radiocarbon dates [@Hsieh2009; @Wang2011]. However, to achieve sample sizes suitable for exploring changes in ornament assemblages over time associated with foreign impacts, we assigned the original excavation levels into three phases: before the European contact, after the European presence, and the presence of large Chinese immigrants.

@@

We made refinements of the original chronology to assign contexts into the pre-European, European, and Chinese periods by following six steps of assigning and evaluating contexts using copies of original excavation records and fieldwork notes. First, for each excavation unit, we assigned the arbitrary excavation layer (mostly 10 cm) to a period based on frequencies of indicators of the European and the Chinese periods. Second, we identified the archaeological context (normally 2-3 layers) that the layer belongs to based on the description of soil colour, texture, and density of potsherds and charcoal. We assumed that a larger population would be reflected by denser distributions of potsherds and charcoal. Third, we checked if there are radiocarbon ages associated with the context as a cross-validation. Since most radiocarbon ages represent long time ranges, we did not consider the ages could determine a period by themselves, but only as a cross-validation. Fourth, we explored any post-depositional issues that might affect the deposition of time indicators to ensure an appropriate assignment based on the fieldwork notes and excavation report. Fifth, we cross-evaluated all adjacent squares to confirm a consistent and continuous context between them. Sixth, we compared our chronology with the six analytic units classified by previous studies to make a final decision [@Hsieh2009; @Wang2011]. More detailed data information is provided in our Supplementary Online Materials. After assigning the contexts into three analytical units, pre-European contact, European presence, and Chinese presence, we assigned ornaments to one of these units according to their find context. We did not use the ornaments themselves as temporal markers because many of the ornaments have been found throughout long time periods in Taiwan. Through those six assigning steps and validation rules, we believe our chronology is reasonable and suitable for discussing the indirect colonial impacts led by trade networks.

The primary archaeological indicators of indirect impacts of Europeans at Kiwulan are the high frequencies of light grey glazed jars, known as "An-ping" jars in China and Taiwan, and large dark brown glazed stoneware jars that were introduced to Taiwan during the early 17th century. Large dark brown glazed stoneware jars may have been made in Southeast Asia, but are frequently found in European shipwrecks from this period as vessels for transporting water, wine or other liquids on long voyages [@Cort2017; @Ketel2011]. The earliest evidence of light grey glazed jars in this region has been found among the cargo of the Spanish shipwreck _San Diego_, which sunk in 1600 CE [@Dizon2016; @Hsieh1995]. Southeast China is assumed to be the origin of the light grey glazed jars, however these are commonly found at sites in Taiwan that were associated with European activities, especially at An-ping in Tainan, where the Zeelandia fort site is located [@Wang2007]. The jar shapes found at Kiwulan are typical of those found elsewhere in VOC sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. We cannot be sure of the exact process that brought them to Kiwulan: they might have been directly imported by Europeans, by Chinese merchants, or by Indigenous groups via regional networks in northern Taiwan. In any case, the high volume of ceramics transported by Europeans, and their high mobility in the shipping trade played an important role in introducing foreign jars to Taiwan.

Those jars were widely distributed across the site and can serve as indicators, together with the radiocarbon dates, to identify the excavation units associated with the pre-European period and the start of European influence at Kiwulan. In addition to stoneware jars as indicators of European presence, around 300 pieces of locally made clay pipes and a few imported pipes were found at Kiwulan. Smoking is likely to have been introduced by Europeans. This custom was widely adopted in many European countries in the 16th century and spread to other regions [@Uccar2019]. We found that the presence of pipe bowls in the archaeological record here is consistent with distributions of glazed jar fragments, which are far more numerous and widespread across the site (n = 1685). It should be noted that these temporal indicators might have been introduced before direct European contact by Chinese traders, and this could result in some uncertainty in identifying the start of European phase. However, the archaeological evidence shows that the layers with abundant trade ceramics match the 17th century according to the excavation report and previous studies [@Hsieh2009; @Wang2011]. Thus, we focus on identifying the contexts with high frequencies of those ceramics as indicators of the early 17th century.
